Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-21 paper.tex
More reshuffling

\subsubsection{Sequence-level preprocessing}
The findings of this study are
based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
up to Aug 22, 2022 and accessible at
% FIXME
doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
if they (1) had ambiguous collection dates; (2) were collected before January
01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
inference, sequences were excluded if they had a ``bad'' quality control status
in any of the four Nextclade columns (``qc.missingData.status'',
``qc.mixedSites.status'', ``qc.frameShifts.status'' and
`qc.stopCodons.status''). These excluded genomes are enriched for low-coverage,
contaminated, or misassembled sequences.

\subsubsection{Site-level preprocessing} Ambiguous nucleotide letters (i.e.,
not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
Problematic bases in the alignments, which had two or more ambiguous letters or
gaps within a distance of seven bases, were masked as missing data following
the protocol in \url{https://github.com/roblanf/sarscov2phylo/}. Masked bases
were ignored when attaching a sample genome to the ARG, but they were
subsequently imputed. Further, 481 problematic sites flagged as prone to
sequencing errors or highly homoplasic were excluded entirely
% Do we need this first URL?
(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).

\subsubsection{Exclusion of ``time travellers''}

\subsection{Inference}
\subsubsection{Overview of sc2ts}
\begin{figure} \centering
\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
\caption{\label{fig:overview_sc2ts}
A schematic of the sc2ts method. Sc2ts reconstructs the genetic relationships
among SARS-CoV-2 genomes by copying samples to all possible ancestors collected
at earlier time points (curved arrows). An ARG is built by attaching new sample
genomes grouped by collection date to an existing ARG rooted at the Wuhan-Hu-1
reference genome, day by day. Each iteration involves three stages: (1)
attachment of new samples to the growing ARG (A, B); (2) reconstruction of
trees relating the samples under each attachment node (C); and parsimony-based
tree topology adjustments (D, E). In the absence of  recombination, sc2ts
infers an ARG that is a single tree relating the samples (A). However, in the
presence of recombination, sc2ts infers an ARG that is encoded as a sequence of
local trees relating segments of the sample genomes (B), inserting a
recombination node to represent a putative recombination event. Additionally,
mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to
make more parsimonious placements of mutations that should be shared or should
not be immediately reverted, respectively.}
\end{figure}

We introduce an ARG inference method, summarised in
Figure~\ref{fig:overview_sc2ts}.
% Note going to rewrite this anyway, so not trying to make it flow.
It uses (1) an efficient HMM implementation of the Li and Stephens (LS) model
\citep{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2)
the reported collection dates of the SARS-CoV-2 samples to determine the order
of genome copying; and (3) several parsimony-based heuristics to locally refine
the topology. In brief, this combined approach uses the likelihood model of LS
to attach sample genomes to existing nodes and to detect recombinants, but
because many of the daily tip samples harbour few, if any, mutational changes,
we use parsimony to resolve their close relationships.

Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by
taking advantage of the reported collection dates to incrementally grow the
ARG. The inference process occurs forward in time, attaching new genomes
aggregated by their collection date (daily batches), from the earliest to the
last collection date. Prior to attaching sample genomes, the ARG is initialised
such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence
of recombination, the ARG is a single tree with no reticulation (Figure 1A);
however, in the presence of recombination, a reticulate ARG is generated, with
variation in tree topology across the genome (Figure 1B). Unlike standard
phylogenetic trees in which internal nodes always represent inferred ancestors,
internal nodes in the sc2ts ARGs represent sequences which can either be
inferred ancestors or known samples.

Each daily batch of genomes is first attached to the growing ARG without any
phylogenetic resolution. In cases where the newly added sequence is inferred to
be a recombinant, a recombination node is inserted; the ``date'' of this node
is arbitrarily set to the average of the collection dates of the genome and the
most recent of its parent nodes. Otherwise, the genomes are attached to the
most closely related ancestral node (an attachment node). This creates
polytomies that are then locally resolved among the descendants of each
attachment node using the Unweighted Pair Group Method with Arithmetic Mean
(UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
This hierarchical clustering algorithm produces a dendrogram relating a set of
sequences, which is treated as an ultrametric binary tree here. Subsequently,
mutations are placed onto the daily tree by maximum parsimony. An issue with
the UPGMA algorithm is that it may create internal nodes supported by no
informative site (i.e., having no mutation immediately ancestral to them). In
an adjustment step, such unsupported internal nodes are removed to
``un-resolve'' the daily tree; the remaining internal nodes, which are
supported by at least one shared mutation, are retained. The ``dates'' of the
non-sample internal nodes are set to the same date as the samples in the daily
batch. The resulting daily tree is grafted onto the attachment node in the
global ARG.

During ARG inference, two types of non-sample nodes are inserted to improve
parsimonious placement of mutations: mutation-coalescing nodes and
pushed-reversion nodes. We call these operations tree adjustments. A
mutation-collapsing node is inserted when a shared mutation between sibling
nodes is found (Figure 1D). An attempt to create mutation-collapsing nodes is
made after grafting a new daily tree onto the global ARG. If shared mutations
exist between two or more sibling nodes, then a parsimonious explanation is to
create a new node that is the parent of those siblings, on which these
mutations occurred. A pushed-reversion node is inserted when an immediate
reversion occurs (Figure 1E), i.e. a mutation that reverts another mutation on
its parent node. Instead of retaining both the mutation and immediate
reversion, only the mutation is kept after a pushed-reversion node is inserted,
thereby reducing the number of mutations on the branches. Note that these
non-sample nodes are used as reconstructed genomes during HMM copying.

Furthermore, during inference, the ARG is automatically encoded, stored, and
updated as a tree sequence using tskit \citep{Kelleher2018-xc}.

\subsubsection{Brief description of the Li and Stephens model} The LS model is
a convenient probabilistic approximation to the coalescent with recombination
\citep{Hudson1983-if}, itself a generative model of a genealogical process with
recombination \citep{Li2003-ib}. Importantly, it can be implemented as an HMM
in which a focal genome is a sequence of nucleotides (observed states) that are
probabilistically emitted as an imperfect mosaic of a set of reference genomes
(hidden states). As such, an efficient HMM machinery can be employed to make
statements about parameters of interest and highlight potential sources of
error in genotyping and sequencing data. The LS model is used in a wide variety
of applications in genomics. For example, it is the engine that drives
statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and
it can be used to estimate mutation rates (e.g., REF), recombination rates
\citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral
sequence data \citep[e.g.][]{Palmer2019-wa}.

@@

\subsubsection{Parameterization of the HMM implementation of Li and Stephens
model} Recombination is inferred using an efficient HMM implementation of the
LS model. Copying paths are estimated from SARS-CoV-2 genomes sampled over
time, with younger genomes copying (i.e. inheriting segments and mutations)
from older genomes. Here, we were able to apply the LS model on millions of
SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of
the LS model that is a core component of the tsinfer software. For this
application, we have parameterised the HMM using a mismatch ratio (MMR), which
tunes how strongly mutations (i.e., no switching of template sequences) are
preferred over recombination (i.e., switching between sequences) during
copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
focal genome more likely spans at least two ancestral genomes (i.e,. the focal
genome is a putative recombinant) than a single ancestral genome. Here, the MMR
was set to three in all the HMM runs (see ``Improvements'' for future potential
development of the HMM).

\subsection{Analysis} \subsubsection{Imputation of Pango lineage for non-sample
nodes} Non-sample nodes inserted during ARG inference had no Pango lineage
assignment from Nextclade. We imputed the Pango lineage of non-sample nodes
using a two-step procedure (Figure 3) to help assess how consistently
breakpoint intervals are estimated by the HMM runs (see below). If the Pango
lineage of the parental node was available and no mutation was immediately
ancestral to the non-sample node, then the parental node’s status was directly
propagated to the non-sample node. Otherwise, the Pango lineage of the
non-sample node was inferred by matching its full set of mutations against the
Pango lineage-defining mutations (based on 90\% consensus of the sequences
analysed) from the COVID-CG website \citep{Chen2021-zc};
\url{https://covidcg.org/}; accessed on November 04, 2022).

\subsubsection{Breakpoint intervals and HMM consistency} When recombinant
ancestry for a given genome is inferred by the LS HMM, a breakpoint is
identified as the last possible position where the copying path must switch.
Thus, the breakpoints used in the ARG represent the rightmost end of the
interval, within which the recombination event could have occurred. To obtain
the leftmost end of this interval, we rerun the HMM using mirrored coordinates
(i.e., from the 3’ end to the 5’ end).

A recombinant sequence was deemed to be HMM-consistent if (1) the same number
of parent nodes and (2) the same set of mutations immediately ancestral to it
were obtained from both the unmirrored and mirrored HMM runs. In the ideal
case, the copying paths of both the HMM runs switch between parent sequences at
the same position(s); in less ideal cases, however, the copying paths switch at
different positions, typically because of a lack of genetic variation between
the positions. The switch positions in the copying paths of both the HMM runs
were used to define the breakpoint interval(s) of an HMM-consistent
recombinant.

>>>>

\subsection{Overview}
Sc2ts is a method for inferring ARGs from densely sampled pandemic-scale data
in real time, in which recombination occurs at a low but significant rate.
The essential idea is to incrementally update the ARG each day
with the sequences collected on that day. These updates are performed by
first finding likely ``copying paths'' for each sequence in the daily
batch to the current ARG, using the ARG-based implementation of
the Li and Stephens model~\citep{Li2003-ib} introduced as
part of \texttt{tsinfer}~\citep{Kelleher2019-ba}. These copying paths
will mostly consist of a new sequence copying from a node in the ARG
with a small number of mutations, and many sequences
from a daily batch often copy from the same ARG node. The second
step is then to ``resolve'' these implied polytomies by using
standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
Such local (greedy) ARG updates inevitably lead to unparsimonious
topologies, and the third step is then to attempt to increase the
overall parsimony of the inference by making some simple topological
updates (Figure~\ref{fig:overview_sc2ts}C and D).

\begin{figure} \centering
\includegraphics[width=0.7\textwidth]{figures/overview_sc2ts.pdf}
\caption{\label{fig:overview_sc2ts}
A schematic of the sc2ts method. Sc2ts reconstructs the genetic relationships
among SARS-CoV-2 genomes by copying samples to all possible ancestors collected
at earlier time points (curved arrows). An ARG is built by attaching new sample
genomes grouped by collection date to an existing ARG rooted at the Wuhan-Hu-1
reference genome, day by day. Each iteration involves three stages: (1)
attachment of new samples to the growing ARG (A, B); (2) reconstruction of
trees relating the samples under each attachment node (C); and parsimony-based
tree topology adjustments (D, E). In the absence of  recombination, sc2ts
infers an ARG that is a single tree relating the samples (A). However, in the
presence of recombination, sc2ts infers an ARG that is encoded as a sequence of
local trees relating segments of the sample genomes (B), inserting a
recombination node to represent a putative recombination event. Additionally,
mutation-collapsing nodes (D) and pushed-reversion nodes (E) are inserted to
make more parsimonious placements of mutations that should be shared or should
not be immediately reverted, respectively.}
\end{figure}

[Quick overview of LS and ARGs/tskit computation here]
As used here, the Li and Stephens model essentially searches among
all sequences in the ARG (both sampled and inferred ancestral nodes)

The details of each of these steps are discussed in subsequent
sections.

% We introduce an ARG inference method, summarised in
% Figure~\ref{fig:overview_sc2ts}.
% % Note going to rewrite this anyway, so not trying to make it flow.
% It uses (1) an efficient HMM implementation of the Li and Stephens (LS) model
% \citep{Li2003-ib} to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2)
% the reported collection dates of the SARS-CoV-2 samples to determine the order
% of genome copying; and (3) several parsimony-based heuristics to locally refine
% the topology. In brief, this combined approach uses the likelihood model of LS
% to attach sample genomes to existing nodes and to detect recombinants, but
% because many of the daily tip samples harbour few, if any, mutational changes,
% we use parsimony to resolve their close relationships.

% Our HMM implementation, sc2ts, builds a reticulated graph for SARS-CoV-2 by
% taking advantage of the reported collection dates to incrementally grow the
% ARG. The inference process occurs forward in time, attaching new genomes
% aggregated by their collection date (daily batches), from the earliest to the
% last collection date. Prior to attaching sample genomes, the ARG is initialised
% such that the Wuhan-Hu-1/2019 reference genome is the root node. In the absence
% of recombination, the ARG is a single tree with no reticulation (Figure 1A);
% however, in the presence of recombination, a reticulate ARG is generated, with
% variation in tree topology across the genome (Figure 1B). Unlike standard
% phylogenetic trees in which internal nodes always represent inferred ancestors,
% internal nodes in the sc2ts ARGs represent sequences which can either be
% inferred ancestors or known samples.

% Each daily batch of genomes is first attached to the growing ARG without any
% phylogenetic resolution. In cases where the newly added sequence is inferred to
% be a recombinant, a recombination node is inserted; the ``date'' of this node
% is arbitrarily set to the average of the collection dates of the genome and the
% most recent of its parent nodes. Otherwise, the genomes are attached to the
% most closely related ancestral node (an attachment node). This creates
% polytomies that are then locally resolved among the descendants of each
% attachment node using the Unweighted Pair Group Method with Arithmetic Mean
% (UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
% This hierarchical clustering algorithm produces a dendrogram relating a set of
% sequences, which is treated as an ultrametric binary tree here. Subsequently,
% mutations are placed onto the daily tree by maximum parsimony. An issue with
% the UPGMA algorithm is that it may create internal nodes supported by no
% informative site (i.e., having no mutation immediately ancestral to them). In
% an adjustment step, such unsupported internal nodes are removed to
% ``un-resolve'' the daily tree; the remaining internal nodes, which are
% supported by at least one shared mutation, are retained. The ``dates'' of the
% non-sample internal nodes are set to the same date as the samples in the daily
% batch. The resulting daily tree is grafted onto the attachment node in the
% global ARG.

% During ARG inference, two types of non-sample nodes are inserted to improve
% parsimonious placement of mutations: mutation-coalescing nodes and
% pushed-reversion nodes. We call these operations tree adjustments. A
% mutation-collapsing node is inserted when a shared mutation between sibling
% nodes is found (Figure 1D). An attempt to create mutation-collapsing nodes is
% made after grafting a new daily tree onto the global ARG. If shared mutations
% exist between two or more sibling nodes, then a parsimonious explanation is to
% create a new node that is the parent of those siblings, on which these
% mutations occurred. A pushed-reversion node is inserted when an immediate
% reversion occurs (Figure 1E), i.e. a mutation that reverts another mutation on
% its parent node. Instead of retaining both the mutation and immediate
% reversion, only the mutation is kept after a pushed-reversion node is inserted,
% thereby reducing the number of mutations on the branches. Note that these
% non-sample nodes are used as reconstructed genomes during HMM copying.

% Furthermore, during inference, the ARG is automatically encoded, stored, and
% updated as a tree sequence using tskit \citep{Kelleher2018-xc}.

\subsection{Ancestral Recombination Graphs}
[Quick overview of ARGs and Tree Sequences/tskit. Refer to the ARG paper for
more discussion and details.]

\subsection{The Li and Stephens model}
[Edit down, cutting the text in the fig caption significantly]

The Li and Stephens (LS) model is
a convenient probabilistic approximation to the coalescent with recombination
\citep{Hudson1983-if}, itself a generative model of a genealogical process with
recombination \citep{Li2003-ib}. Importantly, it can be implemented as an HMM
in which a focal genome is a sequence of nucleotides (observed states) that are
probabilistically emitted as an imperfect mosaic of a set of reference genomes
(hidden states). As such, an efficient HMM machinery can be employed to make
statements about parameters of interest and highlight potential sources of
error in genotyping and sequencing data. The LS model is used in a wide variety
of applications in genomics. For example, it is the engine that drives
statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and
it can be used to estimate mutation rates (e.g., REF), recombination rates
\citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral
sequence data \citep[e.g.][]{Palmer2019-wa}.

@@

Recombination is inferred using an efficient HMM implementation of the
LS model. Copying paths are estimated from SARS-CoV-2 genomes sampled over
time, with younger genomes copying (i.e. inheriting segments and mutations)
from older genomes. Here, we were able to apply the LS model on millions of
SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of
the LS model that is a core component of the tsinfer software. For this
application, we have parameterised the HMM using a mismatch ratio (MMR), which
tunes how strongly mutations (i.e., no switching of template sequences) are
preferred over recombination (i.e., switching between sequences) during
copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
focal genome more likely spans at least two ancestral genomes (i.e,. the focal
genome is a putative recombinant) than a single ancestral genome. Here, the MMR
was set to three in all the HMM runs (see ``Improvements'' for future potential
development of the HMM).

\subsection{Tree inference from clusters of samples}
[Discuss the "local" tree inference, but be careful not to use the term]

\subsection{Parsimony updates}
[Discuss the parsimony tweaks we make to the trees. Get a better name
for the subsection title]

\subsection{Treatment of recombinants}
[Discuss the special treatment we give to recombinants.]
Include the discussion about mirrored coords from here :]

% \subsubsection{Breakpoint intervals and HMM consistency} When recombinant
% ancestry for a given genome is inferred by the LS HMM, a breakpoint is
% identified as the last possible position where the copying path must switch.
% Thus, the breakpoints used in the ARG represent the rightmost end of the
% interval, within which the recombination event could have occurred. To obtain
% the leftmost end of this interval, we rerun the HMM using mirrored coordinates
% (i.e., from the 3’ end to the 5’ end).

% A recombinant sequence was deemed to be HMM-consistent if (1) the same number
% of parent nodes and (2) the same set of mutations immediately ancestral to it
% were obtained from both the unmirrored and mirrored HMM runs. In the ideal
% case, the copying paths of both the HMM runs switch between parent sequences at
% the same position(s); in less ideal cases, however, the copying paths switch at
% different positions, typically because of a lack of genetic variation between
% the positions. The switch positions in the copying paths of both the HMM runs
% were used to define the breakpoint interval(s) of an HMM-consistent
% recombinant.

The findings of this study are
based on sequences and metadata available on GISAID (\url{https://gisaid.org/})
up to Aug 22, 2022 and accessible at
% FIXME
doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
if they (1) had ambiguous collection dates; (2) were collected before January
01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
MN908947.3) using Nextclade v2.3.0~\citep{Aksamentov2021-hj} (dataset tag
2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
inference, sequences were excluded if they had a ``bad'' quality control status
in any of the four Nextclade columns (``qc.missingData.status'',
``qc.mixedSites.status'', ``qc.frameShifts.status'' and
`qc.stopCodons.status''). These excluded genomes are enriched for low-coverage,
contaminated, or misassembled sequences.

Ambiguous nucleotide letters (i.e.,
not A, C, G, T, or a gap) in the pairwise genome alignments were encoded as N.
Problematic bases in the alignments, which had two or more ambiguous letters or
gaps within a distance of seven bases, were masked as missing data following
the protocol in \url{https://github.com/roblanf/sarscov2phylo/}. Masked bases
were ignored when attaching a sample genome to the ARG, but they were
subsequently imputed. Further, 481 problematic sites flagged as prone to
sequencing errors or highly homoplasic were excluded entirely
% Do we need this first URL?
(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).

\subsection{Exclusion of ``time travellers''}
[FIXME I'm pretty sure we don't have two filters here, we just use
maximum sampling delay. Also not sure the MSD acronym is worth it,
we don't use it much.]

\subsection{Imputation of Pango lineage for non-sample nodes}
Non-sample nodes inserted during ARG inference had no Pango lineage
assignment from Nextclade. We imputed the Pango lineage of non-sample nodes
using a two-step procedure (Figure 3) to help assess how consistently
breakpoint intervals are estimated by the HMM runs (see below). If the Pango
lineage of the parental node was available and no mutation was immediately
ancestral to the non-sample node, then the parental node’s status was directly
propagated to the non-sample node. Otherwise, the Pango lineage of the
non-sample node was inferred by matching its full set of mutations against the
Pango lineage-defining mutations (based on 90\% consensus of the sequences
analysed) from the COVID-CG website \citep{Chen2021-zc};
\url{https://covidcg.org/}; accessed on November 04, 2022).
